---
title: "2025 City Tree Analyses"
output: html_document
date: "2025-11-04"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(gghalves)
library(patchwork)
library(scales)
library(ggmap)
library(osmdata)
```



# City Tree Analyses


## Data Wrangle

### Download Seattle Data

'Combined Seattle Trees' data were downloaded on November 4, 2025 from the Seattle GEO Data Portal at  https://data-seattlecitygis.opendata.arcgis.com/datasets/15ca3b333b984fef8fda342ab0bba4d0_0/explore?location=47.632456%2C-122.300026%2C11.90

* Metadata
  + October 3, 2025 at 9:45:22 AM PDT - Info Updated
  + November 4, 2025 at 8:48:14 AM PST - Data Updated
  + August 29, 2023 at 11:22:40 AM PDT - Published Date


### Extract Heat Values

Temperature data was added to Seattle combined trees in Qgis 3.40.5-Bratislava using raster analysis processing tools (out of the box)

* Sample Raster Values
  + Add heat data from each layer to the Seattle tree points
  + Raster Analysis- Sample Raster Values
  + Completed for all trees, for all temperatures (AM,AF,PM)
  + Export .shp and .csv files with enhanced
  
### Import data 

```{r}
#seattle.combined <- read.csv('')
```

