---
title: "2025 City Tree Analyses"
output: html_document
date: "2025-11-04"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(gghalves)
library(patchwork)
library(scales)
library(ggmap)
library(osmdata)
```



# City Tree Analyses


## Data Wrangle

### Download Seattle Data

'Combined Seattle Trees' data were downloaded on November 4, 2025 from the Seattle GEO Data Portal at  https://data-seattlecitygis.opendata.arcgis.com/datasets/15ca3b333b984fef8fda342ab0bba4d0_0/explore?location=47.632456%2C-122.300026%2C11.90

* Metadata
  + October 3, 2025 at 9:45:22 AM PDT - Info Updated
  + November 4, 2025 at 8:48:14 AM PST - Data Updated
  + August 29, 2023 at 11:22:40 AM PDT - Published Date


### Extract Heat Values

Temperature data was added to Seattle combined trees in Qgis 3.40.5-Bratislava using raster analysis processing tools (out of the box)

* Sample Raster Values
  + Add heat data from each layer to the Seattle tree points
  + Raster Analysis- Sample Raster Values
  + Completed for all trees, for all temperatures (AM,AF,PM)
  + Export .shp and .csv files with enhanced
  
### Import data 

```{r}
seattle.combined <- read.csv('https://github.com/jmhulbert/redhot2/raw/refs/heads/main/data/Seattle.Trees.Heat.csv')
```


```{r}
seattle.combined <- seattle.combined %>% filter(kc_af1>0)  %>% mutate(kc.am.f=kc_am1, kc.af.f=kc_af1,kc.pm.f=kc_pm1) %>% mutate(kc.am.c=(((kc.am.f-32)*5)/9),kc.af.c=(((kc.af.f-32)*5)/9),kc.pm.c=(((kc.pm.f-32)*5)/9)) %>% mutate(mean.temp.daily.f=((kc.am.f+kc.af.f+kc.pm.f)/3)) %>% mutate(dist.from.mean.daily.f=mean.temp.daily.f-(mean(mean.temp.daily.f,na.rm=TRUE))) %>% mutate(mean.temp.daily.c=((kc.am.c+kc.af.c+kc.pm.c)/3)) %>% mutate(dist.from.mean.daily.c=mean.temp.daily.c-(mean(mean.temp.daily.c,na.rm=TRUE))) %>% mutate(Area="King County")

```


```{r}
seattle.combined <- seattle.combined %>% mutate(dist.from.mean.am.c = kc.am.c-(mean(kc.am.c,na.rm=TRUE)),dist.from.mean.am.f=kc.am.f-(mean(kc.am.f,na.rm=TRUE))) %>% mutate(dist.from.mean.af.c = kc.af.c-(mean(kc.af.c,na.rm=TRUE)),dist.from.mean.af.f=kc.af.f-(mean(kc.af.f,na.rm=TRUE))) %>% mutate(dist.from.mean.pm.c = kc.pm.c-(mean(kc.pm.c,na.rm=TRUE)),dist.from.mean.pm.f=kc.pm.f-(mean(kc.pm.f,na.rm=TRUE)))
```


```{r}
seattle.streets <- seattle.combined %>% filter(SOURCE_DEP=="Seattle Department of Transportation") %>% mutate(city.space="Street")
seattle.parks <- seattle.combined %>% filter(SOURCE_DEP=="Seattle Parks and Recreation") %>% mutate(city.space="Park")

seattle.combined.back <- rbind(seattle.streets,seattle.parks)
```


```{r}
seattle.wrc <- seattle.combined.back %>% filter(SCIENTIFIC=="Thuja plicata")
```


* King County Dieback Probability Categories (25%)
  + Most Concern (>25%): > 90.29
  + Moderate Concern (5-25%): 85.74 - 90.29
  + Lowest Concern (<5%): < 85.74



```{r}

seattle.most <- seattle.wrc %>% filter(kc.af.f>90.29)%>% mutate(Category25="Most Concern") %>% droplevels()
seattle.mod <- seattle.wrc %>% filter(kc.af.f>85.74 & kc.af.f<90.29) %>% mutate(Category25="Moderate Concern") %>% droplevels()
seattle.low <- seattle.wrc %>% filter(kc.af.f<85.74) %>% mutate(Category25="Lowest Concern") %>% droplevels()

seattle.wrc.cat <- bind_rows(list(seattle.low,seattle.mod,seattle.most))
```


```{r}
ggplot(seattle.wrc.cat,aes(Category25,fill=Category25))+geom_bar(alpha=0.7) +theme_bw() + scale_fill_manual(name="Category",values=c("#7fcdbb","#fe9929","#DDA0DD")) +labs(x="Number of Trees",y="Concern Category") +guides(fill="none")

```


## Seattle Map

```{r}
seabb <- c(left = min(seattle.wrc.cat$lon), 
                              bottom = min(seattle.wrc.cat$lat), 
                              right = max(seattle.wrc.cat$lon), 
                              top = max(seattle.wrc.cat$lat))
```
```{r}
names(seattle.wrc.cat)
```




```{r}
seamap <- get_map(seabb, zoom = 11, scale = 2, maptype="terrain",source="google")
```

```{r}
seattle.map <- ggmap(seamap) + geom_point(data = seattle.combined.cat, aes(x = lon, y = lat,fill=DN_AF1), color = "black",pch=21, size = 3) + theme_minimal() +scale_fill_viridis_c(option = "inferno")+labs(title="Seattle",x="Longitude",y="Latitude",fill="Afternoon\nTemp (F)") +theme(plot.title = element_text(size = 14, hjust = .5,face = "bold.italic"))
```

```{r seattle-map, fig.height=7, fig.width=7}
seattle.map
```


```{r}
seattle.map <- ggmap(seamap) + geom_point(data = seattle.wrc.cat, aes(x = lon, y = lat,fill=Category25), color = "black",pch=21, size = 3) + theme_minimal() +scale_fill_manual(name="Category",values=c("#DDA0DD","#7fcdbb","#fe9929"))+labs(title="Seattle",x="Longitude",y="Latitude",fill="Conservation\nCategory") +theme(plot.title = element_text(size = 14, hjust = .5,face = "bold.italic"))
```
