---
title: "Binomial Distribution Response to Urban Heat - Tacoma"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 3
---

|            |            |            |            |            |
|:----------:|:----------:|:----------:|:----------:|:----------:|
|[Welcome](https://jmhulbert.github.io/redhot2)|[Data](https://jmhulbert.github.io/redhot2/data)|[Analyses](https://jmhulbert.github.io/redhot2/analyses)|[Discussion](https://jmhulbert.github.io/redhot2/discussion)|[Maps](https://jmhulbert.github.io/redhot2/maps)
|             |           |            |            |            |

|            |            |            | Analyses   |            |            |            |
|:----------:|:----------:|:----------:|:----------:|:----------:|:----------:|:----------:|
|[Urban Heat](https://jmhulbert.github.io/redhot2/analyses/heat/)|[All Areas](https://jmhulbert.github.io/redhot2/analyses/heat/all-areas)|[Portland](https://jmhulbert.github.io/redhot2/analyses/heat/portland)|[Tacoma](https://jmhulbert.github.io/redhot2/analyses/heat/tacoma)|[King County](https://jmhulbert.github.io/redhot2/analyses/heat/king)|[iNaturalist](https://jmhulbert.github.io/redhot2/analyses/inaturalist/)|[City Inventories](https://jmhulbert.github.io/redhot2/analyses/city-trees/)|
|             |           |            |            |            |            |            |



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(knitr)
library(kableExtra)
library(gghalves)
library(patchwork)
library(scales)
library(glmmTMB)
library(bbmle)
library(DHARMa)
```

# Tacoma

# Purpose

The purpose of this page is to summarize an investigation of **tree health categories** as a *categorical* response to urban heat (temperatures) for observations from Tacoma, WA.

Response variable: category: *healthy, unhealthy*

*Response variables*
Binary health: healthy, unhealthy
Binary dead top: no top dieback, top dieback
Binary thinning: no thinning, thinning

# Data Wrangling

```{r}
data <- read.csv('https://raw.githubusercontent.com/jmhulbert/redhot2/refs/heads/main/data/urban-data-modified.csv')
```

## Remove dead trees

```{r}
# Remove Dead Trees
data.w.dead <- data 
data <- data %>% filter(field.tree.canopy.symptoms!="Tree is dead") %>% filter(field.dieback.percent<100) %>% droplevels()
```

`r nrow(data.w.dead)-nrow(data)` trees were dead

```{r}
data$binary.tree.canopy.symptoms <- as.factor(data$binary.tree.canopy.symptoms)
levels(data$binary.tree.canopy.symptoms)
```

## Filter to Tacoma Data

```{r}
data <- data %>% filter(Area=="Tacoma") %>% droplevels()
```

`r nrow(data)` trees were included in Tacoma

```{r}
data$top.dieback <- as.factor(data$top.dieback)
data$thinning <- as.factor(data$thinning)
```


## Remove Point Defiance Data (optional)

```{r}
pd.data <- data %>% filter(str_detect(place_guess, "Point Defiance") |str_detect(place_guess, "point defiance")) %>%  mutate(park = str_remove(place_guess, ",.*"))
```

`r nrow(pd.data)` trees were observed in Point Defiance (based on iNat place guess)


```{r}
data.no.pd <- data %>% filter(!str_detect(place_guess, "Point Defiance") & !str_detect(place_guess, "point defiance"))
```

`r nrow(data.no.pd)` trees were observed from throughout Tacoma, not including Point Defiance (based on iNat place guess)


# Visualize Tacoma Data

```{r}
#ggplot(data,aes(binary.tree.canopy.symptoms,tac.af.f,fill=binary.tree.canopy.symptoms))+geom_violin()+coord_flip()+theme_bw()+ scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929"))+guides(fill=FALSE)
```

```{r}
p1 <- ggplot(data,aes(binary.tree.canopy.symptoms,fill=binary.tree.canopy.symptoms))+geom_histogram(stat="count",alpha=0.5)+theme_bw()+coord_flip()+ scale_fill_manual(values=c("#7fcdbb","#fe9929"))+guides(fill=FALSE)
```

```{r}
p2 <- ggplot(data,aes(reclassified.tree.canopy.symptoms,fill=reclassified.tree.canopy.symptoms))+geom_histogram(stat="count",alpha=0.5)+theme_bw()+coord_flip()+ scale_fill_manual(values=c("#fe9929","#7fcdbb","#fe9929","#fe9929"))+guides(fill=FALSE)
```

```{r}
p1 + p2
```

```{r}
#p3 <- ggplot(data,aes(binary.tree.canopy.symptoms)) +facet_wrap(~field.optional...tree.size) +geom_histogram(stat="count")+theme_bw()+coord_flip()
#p3
```

```{r}
#p4 <- ggplot(data,aes(binary.tree.canopy.symptoms)) +facet_wrap(~field.optional...site.location.description) +geom_histogram(stat="count")+theme_bw()+coord_flip()
#p4
```

```{r}
p5 <-ggplot(data,aes(binary.tree.canopy.symptoms,tac.af.f, fill=binary.tree.canopy.symptoms ))+geom_boxplot(alpha=0.5)+theme_bw()+coord_flip() + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929"))+guides(fill=FALSE) +labs(x="Tree Condition",y=NULL)
```

```{r}
p6 <-ggplot(data,aes(tac.af.f,fill=binary.tree.canopy.symptoms))+geom_density(alpha=0.5) + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929")) +theme_bw() +labs(x="Afternoon Temperature") +guides(fill=FALSE)
```

```{r}
p5 / p6
```



## Data without Point Defiance


```{r}
p7 <-ggplot(data.no.pd,aes(binary.tree.canopy.symptoms,tac.af.f, fill=binary.tree.canopy.symptoms))+geom_boxplot(alpha=0.5)+theme_bw()+coord_flip() + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929"))+guides(fill=FALSE) +labs(x="Tree Condition",y=NULL)
```

```{r}
p8 <-ggplot(data.no.pd,aes(tac.af.f,fill=binary.tree.canopy.symptoms))+geom_density(alpha=0.5) + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929")) +theme_bw() +labs(x="Afternoon Temperature") +guides(fill=FALSE)
```

```{r}
p7 / p8
```


## Point Defiance Data Only


```{r}
p9 <-ggplot(pd.data,aes(binary.tree.canopy.symptoms,tac.af.f, fill=binary.tree.canopy.symptoms))+geom_boxplot(alpha=0.5)+theme_bw()+coord_flip() + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929"))+guides(fill=FALSE) +labs(x="Tree Condition",y=NULL)
```

```{r}
p10 <-ggplot(pd.data,aes(tac.af.f,fill=binary.tree.canopy.symptoms))+geom_density(alpha=0.5) + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929")) +theme_bw() +labs(x="Afternoon Temperature") +guides(fill=FALSE)
```


```{r}
p9 / p10
```



# Analyses



## Temperature Time Periods

```{r}
binomial.daily <- glmmTMB(binary.tree.canopy.symptoms ~ mean.temp.daily.f ,family=binomial,data=data)
binomial.am <- glmmTMB(binary.tree.canopy.symptoms ~ tac.am.f ,data=data,family=binomial)
binomial.af <- glmmTMB(binary.tree.canopy.symptoms ~ tac.af.f ,data=data,family=binomial)
binomial.pm <- glmmTMB(binary.tree.canopy.symptoms ~ tac.pm.f,data=data,family=binomial)
```

### Best Fit Comparison

```{r}
AICtab(binomial.daily,binomial.am,binomial.af,binomial.pm)
```

> Overnight heat was the best fit


## Response 1.0: healthy / unhealthy

```{r}
binomial.af <- glmmTMB(binary.tree.canopy.symptoms ~ tac.af.f, data=data,family=binomial)
```

### Summary

```{r}
summary(binomial.af)
```

Tree health is associated with urban heat. The probability of success (tree categorized as unhealthy) increases with urban heat

Unhealthy trees are considered success (1) because it is the second level. The probability of failure (0) is the first level (Healthy).

```{r}
levels(data$binary.tree.canopy.symptoms)
```

The probability of a tree being classified as unhealthy increases with urban temperature. 


### Visualize effects of predictors

First, generate predictions from model

```{r}
# Create a data frame for predictions
new.data <- expand.grid(
  tac.af.f = seq(min(data$tac.af.f), max(data$tac.af.f), length.out = 100)  # keeping predictor2 constant at its mean
)
```

```{r}
new.data$predicted_response <- predict(binomial.af, newdata = new.data, type = "response")
```

```{r}
head(new.data)
```

```{r}
ggplot(new.data, aes(x = tac.af.f, y = predicted_response)) +
  geom_line() +
  labs(x = "Afternoon Temperature", y = "Predicted Response") +
  theme_bw()
```

The probability of tree being classified as unhealthy increases with afternoon temperature




### Compare without Point Defiance Data

```{r}
binomial.af <- glmmTMB(binary.tree.canopy.symptoms ~ tac.af.f, data=data.no.pd,family=binomial)
```

### Summary

```{r}
summary(binomial.af)
```




---

## Response 2.0: healthy / top-dieback

```{r}
data.top <- data %>% filter(reclassified.tree.canopy.symptoms=="Healthy"|reclassified.tree.canopy.symptoms=="Dead Top") %>% droplevels()
```



```{r}
top.dieback.binomial.af <- glmmTMB(top.dieback ~ tac.af.f,data=data.top,family=binomial)
```

### Summary

```{r}
summary(top.dieback.binomial.af)
```


```{r}
levels(data$top.dieback)
```

The probability of a tree having top dieback increases with urban temperature. 

### Visualize effects of predictors

First, generate predictions from model

```{r}
# Create a data frame for predictions
new.data.2 <- expand.grid(
  tac.af.f = seq(min(80), max(150), length.out = 1000)  # keeping predictor2 constant at its mean
)
```


```{r}
head(new.data.2)
```


```{r}
new.data.2$predicted_response <- predict(top.dieback.binomial.af, newdata = new.data.2, type = "response")
```

```{r}
# Create a data frame for predictions
new.data.2.inf <- expand.grid(
  tac.af.f = seq(min(data.top$tac.af.f), max(150), length.out = 1000)  # keeping predictor2 constant at its mean
)
```

```{r}
new.data.2.inf$predicted_response <- predict(top.dieback.binomial.af, newdata = new.data.2.inf, type = "response")
```


```{r}
p <- ggplot(new.data.2.inf, aes(x = tac.af.f, y = predicted_response)) + geom_line() + labs(x = "Afternoon temperature (F)", y = "Dieback Probability") + scale_x_continuous(limits = c(80, 115)) +theme_minimal()

p
```


### Calculate Dieback Probability Categories


```{r}
head(new.data.2[new.data.2$tac.af.f<83.7 & new.data.2$tac.af.f>83,])
```

Tacoma redcedar at 83.12 degrees had a 0.03 probability

```{r}
head(new.data.2[new.data.2$predicted_response>0.024 & new.data.2$predicted_response<0.028,])
```


Tacoma redcedar at 82.42 degrees had a 0.025 probability

```{r}
head(new.data.2[new.data.2$predicted_response>0.048 & new.data.2$predicted_response<0.052,])
```

Tacoma trees at 85.25 had about a 0.05 probability

```{r}
head(new.data.2[new.data.2$predicted_response>0.095,])
```

Tacoma redcedar at about 88.1 degrees had a 0.1 predicted response. 


```{r}
head(new.data.2[new.data.2$predicted_response>0.24 & new.data.2$predicted_response<0.26,])
```

Tacoma redcedar at about 92.36 degrees had a 0.25 predicted response. 


* Tacoma Dieback Probability Categories (25%)
  + Most Concern (>25%): > 92.36
  + Moderate Concern (3-25%): 83.12- 92.36
  + Lowest Concern (<5%): < 83.12


* Tacoma Top-Dieback Probability Categories (10%)
  + Most Concern (>10%): > 88.1
  + Moderate Concern (2.5-10%): 85.25-88.1
  + Lowest Concern (<2.5%): < 85.25 


Convert to Celsius 


85.25 = 29.6
88.1 = 31.2


```{r}
p <- ggplot(new.data.2, aes(x = tac.af.f, y = predicted_response)) + geom_line() + labs(x = "Afternoon temperature (F)", y = "Dieback Probability") + scale_x_continuous(limits = c(80, 92.5),breaks=c(80,82.5,85,87.5,90,92.5)) + scale_y_continuous(limits = c(0, .3)) +theme_minimal()

p2 <- p + annotate(geom="rect",ymin = -Inf, ymax = Inf, 
              xmin = -Inf, xmax = 85.25, fill = alpha('#DDA0DD',0.5)) +
    annotate(geom="rect",ymin = -Inf, ymax = Inf, 
              xmin = 85.25, xmax = 88.1, fill = alpha('#7fcdbb',0.5)) + 
    annotate(geom="rect",ymin = -Inf, ymax = Inf, 
          xmin = 88.1, xmax = Inf, fill = alpha('#fe9929',0.5))
p3 <- p2 + annotate("text", x=82, y=.15, label= "Lowest\nConcern") + annotate("text", x=86.8, y=.15, label= "Moderate\nConcern") + annotate("text", x=91, y=.15, label= "Most\nConcern")
p3
```



---

## Response 3.0: healthy / thinning

```{r}
data.thin <- data %>% filter(reclassified.tree.canopy.symptoms=="Healthy"|reclassified.tree.canopy.symptoms=="Thinning Canopy") %>% droplevels()
```

```{r}
thinning.binomial.af <- glmmTMB(thinning ~ tac.af.f,data=data.thin,family=binomial)
```

### Summary

```{r}
summary(thinning.binomial.af)
```


```{r}
levels(data.thin$thinning)
str(data.thin$thinning)
```

There was no evidence the probability of a tree being classified as thinning is related to afternoon heat.  



### Visualize effects of predictors

First, generate predictions from model

```{r}
# Create a data frame for predictions
new.data.3 <- expand.grid(
  tac.af.f = seq(min(data.thin$tac.af.f), max(data.thin$tac.af.f), length.out = 100),  # keeping predictor2 constant at its mean
  Area = 0  # assuming random effect is zero for prediction
)
```

```{r}
new.data.3$predicted_response <- predict(thinning.binomial.af, newdata = new.data.3, type = "response")
```

```{r}
ggplot(new.data.3, aes(x = tac.af.f, y = predicted_response)) +
  geom_line() +
  labs(x = "Distance from Mean Temperature", y = "Dieback Probability") +
  theme_minimal()
```




