---
title: "iNaturalist Data Summaries"
date: "2025-09-29"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(gghalves)
library(patchwork)
library(scales)
```

# iNATURALIST

# Purpose

The purpose of this page is to provide summaries of the data from the WRDM observations included in these [analyses](https://jmhulbert.github.io/redhot/analyses).

# Data

iNaturalist data were downloaded from the [Western Redcedar Dieback Map](https://www.inaturalist.org/projects/western-redcedar-dieback-map) on 1.13.24.

```{r}
data <- read.csv("./data/urban-data.modified.csv")
```

# Wrangle Data 


## Remove Dead Trees

`r nrow(data)` trees were shared by community scientists in the WRDM project in areas with UHI data between 2020-2023. 

`r nrow(data[data$field.tree.canopy.symptoms=="Tree is dead",])` dead trees were included. 


```{r}
data.w.dead <- data
```


```{r}
data <- data %>% filter(field.tree.canopy.symptoms!="Tree is dead") %>% filter(field.dieback.percent<100) %>% droplevels()
```

```{r}
data$field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.[data$field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.==""] <- "Not sure"
data$field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.[data$field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.=="Unsure"] <- "Not sure"
```


```{r}
hoyt.data <- data.w.dead %>% filter(str_detect(place_guess, "Hoyt") |str_detect(place_guess, "hoyt"))
```

`r nrow(hoyt.data)` trees were added from Hoyt Arboretum.

## Remove Hoyt Trees

```{r}
data <- data %>% filter(!str_detect(place_guess, "Hoyt") & !str_detect(place_guess, "hoyt"))
```

`r nrow(data)` observations were included in analyses after removing dead trees and trees from hoyt.


# Data Summaries

## Observations

### Number of Obs. Per Area

```{r}
data %>% group_by(Area) %>% summarize(n.obs=n())
```






## Community Scientists

### Number of Obs.

```{r}
users <- data %>% group_by(user_login) %>% summarize(n=n()) %>% arrange(desc(n))
users
```

`r nrow(users[users$n==1,])` users only added one observation, but `r nrow(users[users$n>1,])` users added at least two observations, including `r nrow(users[users$n>10,])` who added more than 10, with one person adding as many as 233. 


### Number of Community Scientists

```{r}
data %>% group_by(Area) %>% summarize(n.participants=n_distinct(user_id),n.obs=n())
```


## Tree Health

### Number of Obs. Per Health Category

```{r}
data %>% group_by(reclassified.tree.canopy.symptoms) %>% summarize(n.obs=n())
```

### Number of Obs. per Health Category and Area

```{r}
area.obs.health <- data %>% group_by(Area,reclassified.tree.canopy.symptoms) %>% summarize(n.obs=n())
area.obs.health
```

```{r fig.height=5, fig.width=10}
ggplot(area.obs.health,aes(reclassified.tree.canopy.symptoms,n.obs,fill=reclassified.tree.canopy.symptoms))+facet_wrap(~Area)+geom_col()+coord_flip()+guides(fill="none")+theme_bw()+labs(x="Tree Canopy Symptoms",y="Number of Observations")+geom_text(aes(label=n.obs),nudge_y=20)
```

## Other Factors


### Signs (y/n)


```{r}
data %>% group_by(field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.) %>% summarize(n.obs=n())
```




```{r}
obs.signs <- data %>% group_by(field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.) %>% summarize(n.obs=n())
obs.signs
```


```{r fig.height=5, fig.width=10}
ggplot(obs.signs,aes(field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.,n.obs,fill=field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.))+geom_col()+coord_flip()+guides(fill="none")+theme_bw()+labs(x="Signs or Symptoms Observed",y="Number of Observations")+geom_text(aes(label=n.obs),nudge_y=15)
```

### Factors

```{r}
obs.factors <- data %>% filter(field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.!="No") %>% group_by(field.optional...what..other.factors..were.observed.) %>% summarize(n.obs=n())
obs.factors
```

```{r fig.height=5, fig.width=10}
ggplot(obs.factors,aes(field.optional...what..other.factors..were.observed.,n.obs,fill=field.optional...what..other.factors..were.observed.))+geom_col()+coord_flip()+guides(fill="none")+theme_bw()+labs(x="Other Factor",y="Number of Observations")+geom_text(aes(label=n.obs),nudge_y=15)
```