---
title: "Redhot 2.0 Data Wrangle"
date: "2026-05-20"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message=FALSE, warning=FALSE}
library(tidyverse)
```


|            |            |            |            |            |
|:----------:|:----------:|:----------:|:----------:|:----------:|
|[Welcome](https://jmhulbert.github.io/redhot2)|[Data](https://jmhulbert.github.io/redhot2/data)|[Analyses](https://jmhulbert.github.io/redhot2/analyses)|[Discussion](https://jmhulbert.github.io/redhot2/discussion)|[Maps](https://jmhulbert.github.io/redhot2/maps)
|             |           |            |            |            |

|            |            |            | Analyses   |            |            |            |
|:----------:|:----------:|:----------:|:----------:|:----------:|:----------:|:----------:|
|[Urban Heat](https://jmhulbert.github.io/redhot2/analyses/heat/)|[All Areas](https://jmhulbert.github.io/redhot2/analyses/heat/all-areas)|[Portland](https://jmhulbert.github.io/redhot2/analyses/heat/portland)|[Tacoma](https://jmhulbert.github.io/redhot2/analyses/heat/tacoma)|[King County](https://jmhulbert.github.io/redhot2/analyses/heat/king)|[iNaturalist](https://jmhulbert.github.io/redhot2/analyses/inaturalist/)|[City Inventories](https://jmhulbert.github.io/redhot2/analyses/city-trees/)|
|             |           |            |            |            |            |            |




# Purpose

The purpose of this page is to share the steps taken to wrangle the data for the urban redcedar analyses for reproducability.


# Data

Observations of western redcedar were downloaded from iNaturalist and urban heat data were downloaded from open data portals or provided by contacts in the City of Tacoma, King County (Washington) and Portland. 

## iNaturalist

iNaturalist data were downloaded on **5.13.2026** with the below query. 
    + full data (query: quality_grade=any&identifications=any%projects%5B%5D=western-redcedar-dieback-map)
      + Note 
        + we needed to specify all fields related to place and all fields related to the project
        + we also only included observations fields in core project (not extra fields from other projects present from observations in two projects)
    
      
```{r eval=FALSE}
Query quality_grade=any&identifications=any&projects[]=western-redcedar-dieback-map
Columns id, uuid, observed_on_string, observed_on, time_observed_at, time_zone, user_id, user_login, user_name, created_at, updated_at, quality_grade, license, url, image_url, sound_url, tag_list, description, num_identification_agreements, num_identification_disagreements, captive_cultivated, oauth_application_id, place_guess, latitude, longitude, positional_accuracy, private_place_guess, private_latitude, private_longitude, public_positional_accuracy, geoprivacy, taxon_geoprivacy, coordinates_obscured, positioning_method, positioning_device, place_town_name, place_county_name, place_state_name, place_country_name, place_admin1_name, place_admin2_name, species_guess, scientific_name, common_name, iconic_taxon_name, taxon_id, field:tree+health+condition, field:tree+canopy+symptoms, field:dieback+percent, field:percent+canopy+affected+%28%25%29, field:number+of+additional+unhealthy+trees+%28of+same+species%29+in+area+%28within+sight%29, field:other+factors+-+are+there+signs+or+symptoms+of+insect%2C+diseases%2C+or+other+damage%3F, field:optional+-+what+%27other+factors%27+were+observed%3F, field:optional+-+tree+size, field:optional+-+access+to+water, field:optional+-+estimated+time+spent+to+make+this+observation+%28%23+of+minutes%29, field:tree+number+%28from+google+map%29+if+applicable, field:tree+number+%28from+arboreta+or+other+dataset%29+if+applicable, field:optional+-+signs+of+sap+sucker+feeding%3F, field:notes 
```

## Urban Heat Data

* Urban Heat Island data were obtained for the following locations
  + Portland, OR (downloaded from https://osf.io/eb4tw/ accessed 4.16.22)
  + King County, WA (downloaded from https://osf.io/mz79p/ accessed on 4.16.22)
  + Tacoma, WA (received from City of Tacoma)
  + Pierce County, WA (downloaded from )

> ** Thank you CAPA Strategies for preparing, maintaining and sharing the urban heat data openly !! **

Readme File Notation included with Pierce County 


# Pre-GIS Data Wrangle Methods


```{r}
inat.full <- read.csv("./WRDM-Observations-5.13.26.csv")
```

## Prep for QGIS Data Extracts

Filter data to only include necessary columns 

[1] "id" 
[24] "latitude" 
[25] "longitude" 
[36] "place_town_name" 
[37] "place_county_name" 

```{r}
inat.full.qgis <- inat.full[c(1,24,25,36,37)]
```

```{r}
#write.csv(inat.full.qgis,file="./WRDM-full-5.13.26-qgis.csv")
```


## QGIS Methods

* Import into QGIS
  + Data Source Manager - Delimited Text - Browse to R working directory
  + Add heat data
  
```{r eval=FALSE, include=FALSE}
#python input --- NOTE open python console, then show editor, then copy below code and run
from qgis.core import QgsProject
names = [layer.name() for layer in QgsProject.instance().mapLayers().values()]
print(names)
```

```{r eval=FALSE, include=FALSE}
#python output
['825a_2', '825b_2', '825c_2', 'WRDM-full-5.13.26-qgis', 'af_t_f_ranger', 'am_t_f_ranger', 'clip_tac_af', 'clip_tac_am', 'clip_tac_pm', 'pierce-county-washington_af_temp_f', 'pierce-county-washington_am_temp_f', 'pierce-county-washington_pm_temp_f', 'pm_t_f_ranger']
```
  
```{r eval=FALSE, include=FALSE}
#layers renamed to 
['po_am', 'po_af', 'po_pm', 'WRDM-full-5.13.26-qgis', 'kc_af', 'kc_am', 'tac_af', 'tac_am', 'tac_pm', 'pc_af', 'pc_am', 'pc_pm', 'kc_pm']
```

* Sample Raster Values - Add heat data from each layer to the iNat points
  + Raster Analysis- Sample Raster Values (add geoprocessing toolbox panel)
  
```{r eval=FALSE, include=FALSE}
## Note that I renamed each layer in between runs (first run generated temporary layer 'sampled', second run generated temporary layer 'sampled' - renamed to 'sampled-1' through sampled-12)
QGIS version: 3.44.8-Solothurn
QGIS code revision: 9344f5ed
Qt version: 5.15.18
Python version: 3.12.11
GDAL version: 3.12.0
GEOS version: 3.14.1-CAPI-1.20.5
PROJ version: Rel. 9.7.1, December 1st, 2025
PDAL version: 2.9.3 (git-version: Release)
Algorithm started at: 2026-05-27T07:14:29
Algorithm 'Sample raster values' starting…
Input parameters:
{ 'COLUMN_PREFIX' : 'kc_af', 'INPUT' : 'delimitedtext://file:///Users/professorbabykangaroo/Library/Mobile%20Documents/com~apple~CloudDocs/Core/Server%20Files/redhot2/data/WRDM-full-5.13.26-qgis.csv?type=csv&maxFields=10000&detectTypes=yes&xField=longitude&yField=latitude&crs=EPSG:32610&spatialIndex=no&subsetIndex=no&watchFile=no', 'OUTPUT' : 'TEMPORARY_OUTPUT', 'RASTERCOPY' : '/Users/professorbabykangaroo/Library/Mobile Documents/com~apple~CloudDocs/Core/GIS/Urban Heat GIS data/King County/All Data_CHW_Seattle KingCounty_093020/rasters_chw_seattle king county_120120/af_t_f_ranger.tif' }

Execution completed in 0.13 seconds
Results:
  OUTPUT: Sampled_99eae76e_651a_4d36_9393_f54cec2de14f

Loading resulting layers
Algorithm 'Sample raster values' finished
QGIS version: 3.44.8-Solothurn
QGIS code revision: 9344f5ed
Qt version: 5.15.18
Python version: 3.12.11
GDAL version: 3.12.0
GEOS version: 3.14.1-CAPI-1.20.5
PROJ version: Rel. 9.7.1, December 1st, 2025
PDAL version: 2.9.3 (git-version: Release)
Algorithm started at: 2026-05-27T07:15:04
Algorithm 'Sample raster values' starting…
Input parameters:
{ 'COLUMN_PREFIX' : 'kc_am', 'INPUT' : 'memory://Point?crs=EPSG:4326&field=field_1:integer(0,0)&field=id:integer(0,0)&field=latitude:double(0,0)&field=longitude:double(0,0)&field=place_town_name:string(0,0)&field=place_county_name:string(0,0)&field=kc_af1:double(0,0)&uid={7daa2b9d-6c1b-4b2e-8426-48f813287e0b}', 'OUTPUT' : 'TEMPORARY_OUTPUT', 'RASTERCOPY' : '/Users/professorbabykangaroo/Library/Mobile Documents/com~apple~CloudDocs/Core/GIS/Urban Heat GIS data/King County/All Data_CHW_Seattle KingCounty_093020/rasters_chw_seattle king county_120120/am_t_f_ranger.tif' }

Execution completed in 0.12 seconds
Results:
  OUTPUT: Sampled_903683ac_7cc5_4fe9_b6b8_71d53825430f

Loading resulting layers
Algorithm 'Sample raster values' finished
QGIS version: 3.44.8-Solothurn
QGIS code revision: 9344f5ed
Qt version: 5.15.18
Python version: 3.12.11
GDAL version: 3.12.0
GEOS version: 3.14.1-CAPI-1.20.5
PROJ version: Rel. 9.7.1, December 1st, 2025
PDAL version: 2.9.3 (git-version: Release)
Algorithm started at: 2026-05-27T07:15:34
Algorithm 'Sample raster values' starting…
Input parameters:
{ 'COLUMN_PREFIX' : 'kc_pm', 'INPUT' : 'memory://Point?crs=EPSG:4326&field=field_1:integer(0,0)&field=id:integer(0,0)&field=latitude:double(0,0)&field=longitude:double(0,0)&field=place_town_name:string(0,0)&field=place_county_name:string(0,0)&field=kc_af1:double(0,0)&field=kc_am1:double(0,0)&uid={156e26dd-f361-488a-944a-cc9be7188dd8}', 'OUTPUT' : 'TEMPORARY_OUTPUT', 'RASTERCOPY' : '/Users/professorbabykangaroo/Library/Mobile Documents/com~apple~CloudDocs/Core/GIS/Urban Heat GIS data/King County/All Data_CHW_Seattle KingCounty_093020/rasters_chw_seattle king county_120120/pm_t_f_ranger.tif' }

Execution completed in 0.12 seconds
Results:
  OUTPUT: Sampled_075f57c5_49cf_44c0_b8b3_421f90a1350b

Loading resulting layers
Algorithm 'Sample raster values' finished
QGIS version: 3.44.8-Solothurn
QGIS code revision: 9344f5ed
Qt version: 5.15.18
Python version: 3.12.11
GDAL version: 3.12.0
GEOS version: 3.14.1-CAPI-1.20.5
PROJ version: Rel. 9.7.1, December 1st, 2025
PDAL version: 2.9.3 (git-version: Release)
Algorithm started at: 2026-05-27T07:16:29
Algorithm 'Sample raster values' starting…
Input parameters:
{ 'COLUMN_PREFIX' : 'pc_af', 'INPUT' : 'memory://Point?crs=EPSG:4326&field=field_1:integer(0,0)&field=id:integer(0,0)&field=latitude:double(0,0)&field=longitude:double(0,0)&field=place_town_name:string(0,0)&field=place_county_name:string(0,0)&field=kc_af1:double(0,0)&field=kc_am1:double(0,0)&field=kc_pm1:double(0,0)&uid={7d194a68-6cbd-4579-9586-09100dd51f0f}', 'OUTPUT' : 'TEMPORARY_OUTPUT', 'RASTERCOPY' : '/Users/professorbabykangaroo/Library/Mobile Documents/com~apple~CloudDocs/Core/GIS/Urban Heat GIS data/Pierce County/pierce-county-washington_af_temp_f.tif' }

Execution completed in 0.05 seconds
Results:
  OUTPUT: Sampled_949c45ab_4077_48c9_bb16_f207f3f96794

Loading resulting layers
Algorithm 'Sample raster values' finished
QGIS version: 3.44.8-Solothurn
QGIS code revision: 9344f5ed
Qt version: 5.15.18
Python version: 3.12.11
GDAL version: 3.12.0
GEOS version: 3.14.1-CAPI-1.20.5
PROJ version: Rel. 9.7.1, December 1st, 2025
PDAL version: 2.9.3 (git-version: Release)
Algorithm started at: 2026-05-27T07:16:53
Algorithm 'Sample raster values' starting…
Input parameters:
{ 'COLUMN_PREFIX' : 'pc_am', 'INPUT' : 'memory://Point?crs=EPSG:4326&field=field_1:integer(0,0)&field=id:integer(0,0)&field=latitude:double(0,0)&field=longitude:double(0,0)&field=place_town_name:string(0,0)&field=place_county_name:string(0,0)&field=kc_af1:double(0,0)&field=kc_am1:double(0,0)&field=kc_pm1:double(0,0)&field=pc_af1:double(0,0)&uid={74cf18f7-fe0b-4938-b722-70779ac43334}', 'OUTPUT' : 'TEMPORARY_OUTPUT', 'RASTERCOPY' : '/Users/professorbabykangaroo/Library/Mobile Documents/com~apple~CloudDocs/Core/GIS/Urban Heat GIS data/Pierce County/pierce-county-washington_am_temp_f.tif' }

Execution completed in 0.05 seconds
Results:
  OUTPUT: Sampled_df0abb15_a4ac_4f22_afda_40e939d19b21

Loading resulting layers
Algorithm 'Sample raster values' finished
QGIS version: 3.44.8-Solothurn
QGIS code revision: 9344f5ed
Qt version: 5.15.18
Python version: 3.12.11
GDAL version: 3.12.0
GEOS version: 3.14.1-CAPI-1.20.5
PROJ version: Rel. 9.7.1, December 1st, 2025
PDAL version: 2.9.3 (git-version: Release)
Algorithm started at: 2026-05-27T07:17:20
Algorithm 'Sample raster values' starting…
Input parameters:
{ 'COLUMN_PREFIX' : 'pc_pm', 'INPUT' : 'memory://Point?crs=EPSG:4326&field=field_1:integer(0,0)&field=id:integer(0,0)&field=latitude:double(0,0)&field=longitude:double(0,0)&field=place_town_name:string(0,0)&field=place_county_name:string(0,0)&field=kc_af1:double(0,0)&field=kc_am1:double(0,0)&field=kc_pm1:double(0,0)&field=pc_af1:double(0,0)&field=pc_am1:double(0,0)&uid={e841e5c2-8486-4537-a77e-efd683a5e5ad}', 'OUTPUT' : 'TEMPORARY_OUTPUT', 'RASTERCOPY' : '/Users/professorbabykangaroo/Library/Mobile Documents/com~apple~CloudDocs/Core/GIS/Urban Heat GIS data/Pierce County/pierce-county-washington_pm_temp_f.tif' }

Execution completed in 0.05 seconds
Results:
  OUTPUT: Sampled_0770584e_b859_494c_b52c_16502c4f5ed5

Loading resulting layers
Algorithm 'Sample raster values' finished
QGIS version: 3.44.8-Solothurn
QGIS code revision: 9344f5ed
Qt version: 5.15.18
Python version: 3.12.11
GDAL version: 3.12.0
GEOS version: 3.14.1-CAPI-1.20.5
PROJ version: Rel. 9.7.1, December 1st, 2025
PDAL version: 2.9.3 (git-version: Release)
Algorithm started at: 2026-05-27T07:17:59
Algorithm 'Sample raster values' starting…
Input parameters:
{ 'COLUMN_PREFIX' : 'po_af', 'INPUT' : 'memory://Point?crs=EPSG:4326&field=field_1:integer(0,0)&field=id:integer(0,0)&field=latitude:double(0,0)&field=longitude:double(0,0)&field=place_town_name:string(0,0)&field=place_county_name:string(0,0)&field=kc_af1:double(0,0)&field=kc_am1:double(0,0)&field=kc_pm1:double(0,0)&field=pc_af1:double(0,0)&field=pc_am1:double(0,0)&field=pc_pm1:double(0,0)&uid={bc38586d-25f8-4500-b32a-a334ccdc3f29}', 'OUTPUT' : 'TEMPORARY_OUTPUT', 'RASTERCOPY' : '/Users/professorbabykangaroo/Library/Mobile Documents/com~apple~CloudDocs/Core/GIS/Urban Heat GIS data/Portland/825b_2.tif' }

Execution completed in 0.14 seconds
Results:
  OUTPUT: Sampled_bb2c5487_7f9f_4f26_8f09_019962b58c96

Loading resulting layers
Algorithm 'Sample raster values' finished
QGIS version: 3.44.8-Solothurn
QGIS code revision: 9344f5ed
Qt version: 5.15.18
Python version: 3.12.11
GDAL version: 3.12.0
GEOS version: 3.14.1-CAPI-1.20.5
PROJ version: Rel. 9.7.1, December 1st, 2025
PDAL version: 2.9.3 (git-version: Release)
Algorithm started at: 2026-05-27T07:18:19
Algorithm 'Sample raster values' starting…
Input parameters:
{ 'COLUMN_PREFIX' : 'po_am', 'INPUT' : 'memory://Point?crs=EPSG:4326&field=field_1:integer(0,0)&field=id:integer(0,0)&field=latitude:double(0,0)&field=longitude:double(0,0)&field=place_town_name:string(0,0)&field=place_county_name:string(0,0)&field=kc_af1:double(0,0)&field=kc_am1:double(0,0)&field=kc_pm1:double(0,0)&field=pc_af1:double(0,0)&field=pc_am1:double(0,0)&field=pc_pm1:double(0,0)&field=po_af1:double(0,0)&uid={ef418b3c-340c-4fc0-99d7-b74731644518}', 'OUTPUT' : 'TEMPORARY_OUTPUT', 'RASTERCOPY' : '/Users/professorbabykangaroo/Library/Mobile Documents/com~apple~CloudDocs/Core/GIS/Urban Heat GIS data/Portland/825a_2.tif' }

Execution completed in 0.13 seconds
Results:
  OUTPUT: Sampled_d89e2f26_77c2_468d_b1e5_cd76cd4847e4

Loading resulting layers
Algorithm 'Sample raster values' finished
QGIS version: 3.44.8-Solothurn
QGIS code revision: 9344f5ed
Qt version: 5.15.18
Python version: 3.12.11
GDAL version: 3.12.0
GEOS version: 3.14.1-CAPI-1.20.5
PROJ version: Rel. 9.7.1, December 1st, 2025
PDAL version: 2.9.3 (git-version: Release)
Algorithm started at: 2026-05-27T07:18:47
Algorithm 'Sample raster values' starting…
Input parameters:
{ 'COLUMN_PREFIX' : 'po_pm', 'INPUT' : 'memory://Point?crs=EPSG:4326&field=field_1:integer(0,0)&field=id:integer(0,0)&field=latitude:double(0,0)&field=longitude:double(0,0)&field=place_town_name:string(0,0)&field=place_county_name:string(0,0)&field=kc_af1:double(0,0)&field=kc_am1:double(0,0)&field=kc_pm1:double(0,0)&field=pc_af1:double(0,0)&field=pc_am1:double(0,0)&field=pc_pm1:double(0,0)&field=po_af1:double(0,0)&field=po_am1:double(0,0)&uid={f6a055d1-6e22-4c11-b708-949a79d2661c}', 'OUTPUT' : 'TEMPORARY_OUTPUT', 'RASTERCOPY' : '/Users/professorbabykangaroo/Library/Mobile Documents/com~apple~CloudDocs/Core/GIS/Urban Heat GIS data/Portland/825c_2.tif' }

Execution completed in 0.13 seconds
Results:
  OUTPUT: Sampled_68408d81_336e_414d_9e3f_351e7d6048e9

Loading resulting layers
Algorithm 'Sample raster values' finished
QGIS version: 3.44.8-Solothurn
QGIS code revision: 9344f5ed
Qt version: 5.15.18
Python version: 3.12.11
GDAL version: 3.12.0
GEOS version: 3.14.1-CAPI-1.20.5
PROJ version: Rel. 9.7.1, December 1st, 2025
PDAL version: 2.9.3 (git-version: Release)
Algorithm started at: 2026-05-27T07:19:09
Algorithm 'Sample raster values' starting…
Input parameters:
{ 'COLUMN_PREFIX' : 'tac_af', 'INPUT' : 'memory://Point?crs=EPSG:4326&field=field_1:integer(0,0)&field=id:integer(0,0)&field=latitude:double(0,0)&field=longitude:double(0,0)&field=place_town_name:string(0,0)&field=place_county_name:string(0,0)&field=kc_af1:double(0,0)&field=kc_am1:double(0,0)&field=kc_pm1:double(0,0)&field=pc_af1:double(0,0)&field=pc_am1:double(0,0)&field=pc_pm1:double(0,0)&field=po_af1:double(0,0)&field=po_am1:double(0,0)&field=po_pm1:double(0,0)&uid={4c759164-6d13-4881-9995-2011e06ff6a2}', 'OUTPUT' : 'TEMPORARY_OUTPUT', 'RASTERCOPY' : '/Users/professorbabykangaroo/Library/Mobile Documents/com~apple~CloudDocs/Core/GIS/Urban Heat GIS data/Tacoma/tac/clip_tac_af.tif' }

Execution completed in 0.05 seconds
Results:
  OUTPUT: Sampled_78e3fc06_f49a_495c_97d6_ff219505258f

Loading resulting layers
Algorithm 'Sample raster values' finished
QGIS version: 3.44.8-Solothurn
QGIS code revision: 9344f5ed
Qt version: 5.15.18
Python version: 3.12.11
GDAL version: 3.12.0
GEOS version: 3.14.1-CAPI-1.20.5
PROJ version: Rel. 9.7.1, December 1st, 2025
PDAL version: 2.9.3 (git-version: Release)
Algorithm started at: 2026-05-27T07:19:29
Algorithm 'Sample raster values' starting…
Input parameters:
{ 'COLUMN_PREFIX' : 'tac_am', 'INPUT' : 'memory://Point?crs=EPSG:4326&field=field_1:integer(0,0)&field=id:integer(0,0)&field=latitude:double(0,0)&field=longitude:double(0,0)&field=place_town_name:string(0,0)&field=place_county_name:string(0,0)&field=kc_af1:double(0,0)&field=kc_am1:double(0,0)&field=kc_pm1:double(0,0)&field=pc_af1:double(0,0)&field=pc_am1:double(0,0)&field=pc_pm1:double(0,0)&field=po_af1:double(0,0)&field=po_am1:double(0,0)&field=po_pm1:double(0,0)&field=tac_af1:double(0,0)&uid={de2b674d-d920-46bf-82c2-602b4eadbad9}', 'OUTPUT' : 'TEMPORARY_OUTPUT', 'RASTERCOPY' : '/Users/professorbabykangaroo/Library/Mobile Documents/com~apple~CloudDocs/Core/GIS/Urban Heat GIS data/Tacoma/tac/clip_tac_am.tif' }

Execution completed in 0.05 seconds
Results:
  OUTPUT: Sampled_4cebe462_6972_4a14_b9a2_8f799f6441c4

Loading resulting layers
Algorithm 'Sample raster values' finished
QGIS version: 3.44.8-Solothurn
QGIS code revision: 9344f5ed
Qt version: 5.15.18
Python version: 3.12.11
GDAL version: 3.12.0
GEOS version: 3.14.1-CAPI-1.20.5
PROJ version: Rel. 9.7.1, December 1st, 2025
PDAL version: 2.9.3 (git-version: Release)
Algorithm started at: 2026-05-27T07:19:57
Algorithm 'Sample raster values' starting…
Input parameters:
{ 'COLUMN_PREFIX' : 'tac_pm', 'INPUT' : 'memory://Point?crs=EPSG:4326&field=field_1:integer(0,0)&field=id:integer(0,0)&field=latitude:double(0,0)&field=longitude:double(0,0)&field=place_town_name:string(0,0)&field=place_county_name:string(0,0)&field=kc_af1:double(0,0)&field=kc_am1:double(0,0)&field=kc_pm1:double(0,0)&field=pc_af1:double(0,0)&field=pc_am1:double(0,0)&field=pc_pm1:double(0,0)&field=po_af1:double(0,0)&field=po_am1:double(0,0)&field=po_pm1:double(0,0)&field=tac_af1:double(0,0)&field=tac_am1:double(0,0)&uid={298c93b9-246b-4941-93d1-d0c7ab38c3a2}', 'OUTPUT' : 'TEMPORARY_OUTPUT', 'RASTERCOPY' : '/Users/professorbabykangaroo/Library/Mobile Documents/com~apple~CloudDocs/Core/GIS/Urban Heat GIS data/Tacoma/tac/clip_tac_pm.tif' }

Execution completed in 0.05 seconds
Results:
  OUTPUT: Sampled_2d98258d_74ef_43f1_b00c_cffbe8f8f58a

Loading resulting layers
Algorithm 'Sample raster values' finished

# in the end i had one layer with temperature values for all times/cities/counties for points within those areas
```
  


# Post-GIS Data Wrangle

## Re-import data after QGIS methods

```{r}
inat.full.heat <- read.csv("./WRDM-full-5.13.26-qgis-w-heat.csv")
```

## Join heat data with iNat data

```{r}
redhot <- left_join(inat.full.heat,inat.full,by="id")
```

## Subset based on counties
  
```{r}
tac.wrc <- redhot %>% filter(tac_af1>0) %>% mutate(tac.am.c=tac_am1, tac.af.c=tac_af1,tac.pm.c=tac_pm1)
pc.wrc <- redhot %>% filter(pc_af1>0  & is.na(tac_af1) & is.na(kc_af1)) %>% mutate(pc.am.f=pc_am1, pc.af.f=pc_af1,pc.pm.f=pc_pm1)
kc.wrc <- redhot %>% filter(kc_af1>0 & is.na(tac_af1) & is.na(pc_af1))  %>% mutate(kc.am.f=kc_am1, kc.af.f=kc_af1,kc.pm.f=kc_pm1)
pl.wrc <- redhot %>% filter(po_af1>0) %>% mutate(pl.am.c=po_am1, pl.af.c=po_af1,pl.pm.c=po_pm1)
```

* Number of observations
  + Tacoma - 1739
  + Pierce County - 29 (21 after exluding trees overlapping with tacoma and king county)
  + King County - 601 (589 after excluding trees overlapping with pierce county and tacoma)
  + Portland - 677
  
> Note Tacoma trees overlapping data from king county or pierce county were included in Tacoma dataset instead of king county or pierce county datasets.  King county and pierce county datasets do not include Trees with tacoma data. King county data does not included trees with pierce county data.  
  
## Mutate Data (per area)

> Note we needed both f and c temperatures for each area. Tacoma and Portland data were 'sample with raster values' from a dataset in C and King County and Pierce County were sampled from dataset in F.

### daily means

```{r}
tac.wrc <- tac.wrc %>% mutate(tac.am.f=((tac.am.c*1.8)+32),tac.af.f=((tac.af.c*1.8)+32),tac.pm.f=((tac.pm.c*1.8)+32)) %>% mutate(mean.temp.daily.f=((tac.am.f+tac.af.f+tac.pm.f)/3)) %>% mutate(dist.from.mean.daily.f=mean.temp.daily.f-(mean(mean.temp.daily.f,na.rm=TRUE))) %>% mutate(mean.temp.daily.c=((tac.am.c+tac.af.c+tac.pm.c)/3)) %>% mutate(dist.from.mean.daily.c=mean.temp.daily.c-(mean(mean.temp.daily.c,na.rm=TRUE))) %>% mutate(Area="Tacoma")

pc.wrc <- pc.wrc %>% mutate(pc.am.c=(((pc.am.f-32)*5)/9),pc.af.c=(((pc.af.f-32)*5)/9),pc.pm.c=(((pc.pm.f-32)*5)/9)) %>% mutate(mean.temp.daily.f=((pc.am.f+pc.af.f+pc.pm.f)/3)) %>% mutate(dist.from.mean.daily.f=mean.temp.daily.f-(mean(mean.temp.daily.f,na.rm=TRUE))) %>% mutate(mean.temp.daily.c=((pc.am.c+pc.af.c+pc.pm.c)/3)) %>% mutate(dist.from.mean.daily.c=mean.temp.daily.c-(mean(mean.temp.daily.c,na.rm=TRUE))) %>% mutate(Area="Pierce County")

kc.wrc <- kc.wrc %>% mutate(kc.am.c=(((kc.am.f-32)*5)/9),kc.af.c=(((kc.af.f-32)*5)/9),kc.pm.c=(((kc.pm.f-32)*5)/9)) %>% mutate(mean.temp.daily.f=((kc.am.f+kc.af.f+kc.pm.f)/3)) %>% mutate(dist.from.mean.daily.f=mean.temp.daily.f-(mean(mean.temp.daily.f,na.rm=TRUE))) %>% mutate(mean.temp.daily.c=((kc.am.c+kc.af.c+kc.pm.c)/3)) %>% mutate(dist.from.mean.daily.c=mean.temp.daily.c-(mean(mean.temp.daily.c,na.rm=TRUE))) %>% mutate(Area="King County")

pl.wrc <- pl.wrc %>% mutate(pl.am.f=((pl.am.c*1.8)+32),pl.af.f=((pl.af.c*1.8)+32),pl.pm.f=((pl.pm.c*1.8)+32)) %>% mutate(mean.temp.daily.f=((pl.am.f+pl.af.f+pl.pm.f)/3)) %>% mutate(dist.from.mean.daily.f=mean.temp.daily.f-(mean(mean.temp.daily.f,na.rm=TRUE))) %>% mutate(mean.temp.daily.c=((pl.am.c+pl.af.c+pl.pm.c)/3)) %>% mutate(dist.from.mean.daily.c=mean.temp.daily.c-(mean(mean.temp.daily.c,na.rm=TRUE))) %>% mutate(Area="Portland")
```


### morning means

```{r}
tac.wrc <- tac.wrc %>% mutate(dist.from.mean.am.c = tac.am.c-(mean(tac.am.c,na.rm=TRUE)),dist.from.mean.am.f=tac.am.f-(mean(tac.am.f,na.rm=TRUE)))
pc.wrc <- pc.wrc %>% mutate(dist.from.mean.am.c = pc.am.c-(mean(pc.am.c,na.rm=TRUE)),dist.from.mean.am.f=pc.am.f-(mean(pc.am.f,na.rm=TRUE)))
kc.wrc <- kc.wrc %>% mutate(dist.from.mean.am.c = kc.am.c-(mean(kc.am.c,na.rm=TRUE)),dist.from.mean.am.f=kc.am.f-(mean(kc.am.f,na.rm=TRUE)))
pl.wrc <- pl.wrc %>%  mutate(dist.from.mean.am.c = pl.am.c-(mean(pl.am.c,na.rm=TRUE)),dist.from.mean.am.f=pl.am.f-(mean(pl.am.f,na.rm=TRUE)))
```

### afternoon means

```{r}
tac.wrc <- tac.wrc %>% mutate(dist.from.mean.af.c = tac.af.c-(mean(tac.af.c,na.rm=TRUE)),dist.from.mean.af.f=tac.af.f-(mean(tac.af.f,na.rm=TRUE)))
pc.wrc <- pc.wrc %>% mutate(dist.from.mean.af.c = pc.af.c-(mean(pc.af.c,na.rm=TRUE)),dist.from.mean.af.f=pc.af.f-(mean(pc.af.f,na.rm=TRUE)))
kc.wrc <- kc.wrc %>% mutate(dist.from.mean.af.c = kc.af.c-(mean(kc.af.c,na.rm=TRUE)),dist.from.mean.af.f=kc.af.f-(mean(kc.af.f,na.rm=TRUE)))
pl.wrc <- pl.wrc %>%  mutate(dist.from.mean.af.c = pl.af.c-(mean(pl.af.c,na.rm=TRUE)),dist.from.mean.af.f=pl.af.f-(mean(pl.af.f,na.rm=TRUE)))
```


### evening means

```{r}
tac.wrc <- tac.wrc %>% mutate(dist.from.mean.pm.c = tac.pm.c-(mean(tac.pm.c,na.rm=TRUE)),dist.from.mean.pm.f=tac.pm.f-(mean(tac.pm.f,na.rm=TRUE)))
pc.wrc <- pc.wrc %>% mutate(dist.from.mean.pm.c = pc.pm.c-(mean(pc.pm.c,na.rm=TRUE)),dist.from.mean.pm.f=pc.pm.f-(mean(pc.pm.f,na.rm=TRUE)))
kc.wrc <- kc.wrc %>% mutate(dist.from.mean.pm.c = kc.pm.c-(mean(kc.pm.c,na.rm=TRUE)),dist.from.mean.pm.f=kc.pm.f-(mean(kc.pm.f,na.rm=TRUE)))
pl.wrc <- pl.wrc %>%  mutate(dist.from.mean.pm.c = pl.pm.c-(mean(pl.pm.c,na.rm=TRUE)),dist.from.mean.pm.f=pl.pm.f-(mean(pl.pm.f,na.rm=TRUE)))
```


## Merge Data (merge areas)

```{r}
data <- bind_rows(tac.wrc,pc.wrc,kc.wrc,pl.wrc)
```


## Clean Data

Some of the iNat project questions changed since it was created so some we need to adjust the answers to be more consistent throughout the project. 

### Clean iNaturalist Fields

```{r}
data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. <- as.factor(data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
data$field.optional...what..other.factors..were.observed. <- as.factor(data$field.optional...what..other.factors..were.observed.)
data$field.tree.canopy.symptoms <- as.factor(data$field.tree.canopy.symptoms)
#data$field.optional...slope.position <- as.factor(data$field.optional...slope.position) #Some reason field was not available in export list of field in iNat?
#data$field.optional...site.type <- as.factor(data$field.optional...site.type)  #Some reason field was not available in export list of field in iNat?
#data$field.optional...site.location.description  <- as.factor(data$field.optional...site.location.description ) #Some reason field was not available in export list of field in iNat?
data$field.optional...tree.size <-as.factor(data$field.optional...tree.size)
```


```{r}
data$field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.[data$field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.==""] <- "Not sure"
data$field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.[data$field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage.=="Unsure"] <- "Not sure"
```


```{r}
data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="4"] <- "4-6"
data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="5"] <- "4-6"
data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="2"] <- "2-3"
```


```{r}
data$field.tree.canopy.symptoms[data$field.tree.canopy.symptoms=="Multiple Symptoms"] <-"Multiple Symptoms (please list in Notes)"
data$field.tree.canopy.symptoms[data$field.tree.canopy.symptoms=="multiple symptoms"] <-"Multiple Symptoms (please list in Notes)"
data$field.tree.canopy.symptoms[data$field.tree.canopy.symptoms=="thinning foliage"] <-"Thinning Canopy"
data$field.tree.canopy.symptoms[data$field.tree.canopy.symptoms=="healthy"] <-"Healthy"
data$field.tree.canopy.symptoms[data$field.tree.canopy.symptoms=="dead top"] <-"Old Dead Top (needles already gone)"
```


```{r}
data$field.optional...what..other.factors..were.observed.[data$field.optional...what..other.factors..were.observed.=="Fungal Activitiy (mycelial fans, mushrooms at base, or conks on trunk)"] <-"Fungal Activitiy (mycelial fans, bleeding cankers, mushrooms at base, or conks on trunk)"
data$field.optional...what..other.factors..were.observed.[data$field.optional...what..other.factors..were.observed.=="Needle disease (dieback, checking, blight, etc.)"] <- "Needle or leaf disease (dieback, checking, blight, etc.)"
```


```{r}
#data$field.optional...slope.position[data$field.optional...slope.position=="Upper 1/3rd of a slope"] <-"Top of slope" #Some reason field was not available in export list of field in iNat?
```


```{r}
#Some reason field was not available in export list of field in iNat?
#data$field.optional...site.type[data$field.optional...site.type=="Urban Natural"] <-"Urban"
#data$field.optional...site.type[data$field.optional...site.type=="Urban Landscaped"] <-"Urban"
#data$field.optional...site.type[data$field.optional...site.type=="Suburban Natural"] <-"Suburban"
#data$field.optional...site.type[data$field.optional...site.type=="Suburban Lanscaped"] <-"Suburban"
#data$field.optional...site.type[data$field.optional...site.type=="Natural Forest"] <-"Rural"
```

```{r}
levels(data$field.optional...tree.size)
```


```{r}
data$field.optional...tree.size[data$field.optional...tree.size=="Large"] <- "Large (too big to wrap arms around trunk)"
data$field.optional...tree.size[data$field.optional...tree.size=="Medium"] <- "Medium (can wrap arms around trunk)"
data$field.optional...tree.size[data$field.optional...tree.size=="Small"] <- "Small (can wrap hands around trunk)"
data$field.optional...tree.size[data$field.optional...tree.size=="Very Large"] <- "Very Large (would take many people to wrap arms around trunk)"
data$field.optional...tree.size[data$field.optional...tree.size=="Very small (can wrap a single hand around stem)"] <- "Very small (can wrap a single hand around stem)"
```


```{r}
#Some reason field was not available in export list of field in iNat?
#data$field.optional...site.location.description [data$field.optional...site.location.description =="Yard or open park grounds"] <- "Urban yard or open park grounds" 
```

### Reclassify co-factors

```{r}
data$field.percent.canopy.affected.... <- as.factor(data$field.percent.canopy.affected....)
data$field.percent.canopy.affected....[data$field.percent.canopy.affected....=="1-25% of the crown is unhealthy"] <- "1-29% of the canopy is unhealthy"
data$field.percent.canopy.affected....[data$field.percent.canopy.affected....=="Healthy (0%)"] <- "Healthy, no dieback(0%)"
data$field.percent.canopy.affected....[data$field.percent.canopy.affected....=="Healthy (0% is unhealthy)"] <- "Healthy, no dieback(0%)"
data$field.percent.canopy.affected....[data$field.percent.canopy.affected....=="more than 75% of the crown is unhealthy"] <- "60-99% of the canopy is unhealthy"
data <- data %>% droplevels()

data$Percent.Dieback.Modified[data$field.percent.canopy.affected....=="Healthy, no dieback(0%)"] <- 0
data$Percent.Dieback.Modified[data$field.percent.canopy.affected....=="1-29% of the canopy is unhealthy"] <- 1
data$Percent.Dieback.Modified[data$field.percent.canopy.affected....=="30-59% of the canopy is unhealthy"] <- 30
data$Percent.Dieback.Modified[data$field.percent.canopy.affected....=="60-99% of the canopy is unhealthy"] <- 60
data$Percent.Dieback.Modified[data$field.percent.canopy.affected....=="tree is dead"] <- 100
#data$field.dieback.percent[is.na(data$field.dieback.percent)] <- data$Percent.Dieback.Modified

data$field.percent.canopy.affected....[data$Percent.Dieback.Modified==100] <- "tree is dead" #there were a couple healthy trees with 100 dieback somehow..

has.dieback.percent <- data %>% filter(field.dieback.percent!="")
does.not.have.dieback.percent <- data %>% filter(field.dieback.percent=="") 

has.dieback.percent$user.estimated.dieback <- "Yes"
does.not.have.dieback.percent$user.estimated.dieback <- "No"

does.not.have.dieback.percent$field.dieback.percent <- does.not.have.dieback.percent$Percent.Dieback.Modified

data <- rbind(has.dieback.percent,does.not.have.dieback.percent)

data$field.dieback.percent[data$field.dieback.percent<0] <- 0 # not sure why but there was one -2 percent dieback value

```



```{r}
data <- data %>% mutate(tree.size.simplified=field.optional...tree.size) 
```

```{r}
tree_size_level_key <- c("Very Small (can wrap a single hand around stem)" = "Small", "Small (can wrap hands around trunk)" = "Small", "Medium (can wrap arms around trunk)" = "Medium", "Large (too big to wrap arms around trunk)" = "Large", "Very Large (would take many people to wrap arms around trunk)" = "Large","Other"="Other","No selection"="No Selection")

data$tree.size.simplified <- recode_factor(data$tree.size.simplified, !!!tree_size_level_key)
data$tree.size.simplified <- as.factor(data$tree.size.simplified)
```


### Reclassify response category variables

```{r}
data <- data %>% filter(field.tree.canopy.symptoms!="Candelabra top or very old spike top (old growth)") %>% mutate(binary.tree.canopy.symptoms=field.tree.canopy.symptoms) %>% mutate(ordinal.tree.canopy.symptoms=field.tree.canopy.symptoms) %>% mutate(reclassified.tree.canopy.symptoms=field.tree.canopy.symptoms) %>% droplevels()
```


```{r}
binary_level_key <- c("Healthy" = "Healthy", "Thinning Canopy" = "Unhealthy", "New Dead Top (red or brown needles still attached)" = "Unhealthy", "Old Dead Top (needles already gone)" = "Unhealthy", "Tree is dead" = "Unhealthy", "Multiple Symptoms (please list in Notes)" = "Unhealthy", "Extra Cone Crop" = "Unhealthy", "Browning Canopy" = "Unhealthy","Branch Dieback or 'Flagging'" = "Unhealthy", "Other (please describe in Notes)" = "Unhealthy", "Yellowing Canopy" = "Unhealthy")

data$binary.tree.canopy.symptoms <- recode_factor(data$binary.tree.canopy.symptoms, !!!binary_level_key)
#levels(binary$field.tree.canopy.symptoms)
data$binary.tree.canopy.symptoms <- as.factor(data$binary.tree.canopy.symptoms)
```


```{r}
ordinal_level_key <- c("Healthy" = "Healthy", "Thinning Canopy" = "Unhealthy", "New Dead Top (red or brown needles still attached)" = "Unhealthy", "Old Dead Top (needles already gone)" = "Unhealthy", "Tree is dead" = "Dead", "Multiple Symptoms (please list in Notes)" = "Unhealthy", "Extra Cone Crop" = "Unhealthy", "Browning Canopy" = "Unhealthy","Branch Dieback or 'Flagging'" = "Unhealthy", "Other (please describe in Notes)" = "Unhealthy", "Yellowing Canopy" = "Unhealthy")

data$ordinal.tree.canopy.symptoms <- recode_factor(data$ordinal.tree.canopy.symptoms, !!!ordinal_level_key)
#levels(binary$field.tree.canopy.symptoms)
data$ordinal.tree.canopy.symptoms <- as.factor(data$ordinal.tree.canopy.symptoms)
```


```{r}
reclassified_level_key <- c("Healthy" = "Healthy", "Thinning Canopy" = "Thinning Canopy", "New Dead Top (red or brown needles still attached)" = "Dead Top", "Old Dead Top (needles already gone)" = "Dead Top", "Tree is dead" = "Tree is Dead", "Multiple Symptoms (please list in Notes)" = "Other", "Extra Cone Crop" = "Other", "Browning Canopy" = "Other","Branch Dieback or 'Flagging'" = "Other", "Other (please describe in Notes)" = "Other", "Yellowing Canopy" = "Other")

data$reclassified.tree.canopy.symptoms <- recode_factor(data$reclassified.tree.canopy.symptoms, !!!reclassified_level_key)
#levels(binary$field.tree.canopy.symptoms)
data$reclassified.tree.canopy.symptoms <- as.factor(data$reclassified.tree.canopy.symptoms)
```

Create Binary response for dead top

```{r}
data$top.dieback[data$reclassified.tree.canopy.symptoms=="Dead Top"] <- "Yes"
data$top.dieback[data$reclassified.tree.canopy.symptoms!="Dead Top"] <- "No"
data$top.dieback <- as.factor(data$top.dieback)
levels(data$top.dieback)
summary(data$top.dieback)
```

Create Binary response for thinning

```{r}
data$thinning[data$reclassified.tree.canopy.symptoms=="Thinning Canopy"] <- "Yes"
data$thinning[data$reclassified.tree.canopy.symptoms!="Thinning Canopy"] <- "No"
data$thinning <- as.factor(data$thinning)
levels(data$thinning)
summary(data$thinning)
```

Create Binary response for dead

```{r}
data$dead[data$reclassified.tree.canopy.symptoms=="Tree is Dead"] <- "Yes"
data$dead[data$reclassified.tree.canopy.symptoms!="Tree is Dead"] <- "No"
data$dead <- as.factor(data$dead)
levels(data$dead)
summary(data$dead)
```


## Mutate Merged Data

### Convert Percent to Proportion



```{r}
data$field.dieback.percent <- as.numeric(data$field.dieback.percent)
data <- data %>% mutate(field.dieback.prop = (field.dieback.percent/100))
```


## Filter Merged Data

### Remove other species

A few observations were not identified to species or were determined to be other species by the iNat community. Two observations were "Thuja" and were kept in the dataset. 

```{r}
levels(as.factor(data$scientific_name))
```

```{r}
other.species <- data %>% filter(scientific_name!="Thuja plicata") %>% filter(scientific_name!="Thuja") %>% droplevels()
data <- data %>% filter(scientific_name=="Thuja plicata" | scientific_name=="Thuja") %>% droplevels()
```

> **`r length(other.species$scientific_name)`** observations were identified as other species by the iNaturalist community or were not identified to species. 


### Check for outliers

Check oservations to see if any are more than +/- 5 degrees C from mean.

```{r}
ggplot(data,aes(dist.from.mean.af.c,field.dieback.percent))+geom_point()+theme_bw()
```

### Remove innacurate locations
```{r}
ggplot(data,aes(y=latitude.y,x=longitude.y,size=positional_accuracy)) +geom_point() +theme_bw()
```

Data were filtered to those less than 4000 m positional accuracy to limit incorrect temperatures because climate NA is based on 4km grid cells.

```{r}
data <- data %>% filter(positional_accuracy<=4000)
```

```{r}
ggplot(data,aes(y=latitude.y,x=longitude.y,size=positional_accuracy)) +geom_point() +theme_bw()
```


## Export Modified Data

```{r}
#write.csv(data,file="./urban-data-modified.csv")
```

[urban-data-modified.csv](./urban-data-modified.csv)

Please use this data in analyses. Please make any changes or corrections to the data in this R markdown so everyone is using the same dataset in the analyses.



