---
title: "Binomial Distribution Response to Urban Heat - King County"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 3
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(corrplot)
library(knitr)
library(kableExtra)
library(gghalves)
library(patchwork)
library(scales)
library(glmmTMB)
library(bbmle)
library(DHARMa)
```

# King County

# Purpose

The purpose of this page is to summarize an investigation of **tree health categories** as a *categorical* response to urban heat (temperatures) for observations from King County, WA.

Response variable: category: *healthy, unhealthy*

*Response variables*
Binary health: healthy, unhealthy
Binary dead top: no top dieback, top dieback
Binary thinning: no thinning, thinning

# Data Wrangling

```{r}
data <- read.csv('https://raw.githubusercontent.com/jmhulbert/redhot2/refs/heads/main/data/urban-data-modified.csv')
```

## Remove dead trees

```{r}
# Remove Dead Trees
data.w.dead <- data 
data <- data %>% filter(field.tree.canopy.symptoms!="Tree is dead") %>% filter(field.dieback.percent<100) %>% droplevels()
```

`r nrow(data.w.dead)-nrow(data)` trees were dead

```{r}
data$binary.tree.canopy.symptoms <- as.factor(data$binary.tree.canopy.symptoms)
levels(data$binary.tree.canopy.symptoms)
```

## Filter to King County Data

```{r}
data <- data %>% filter(Area=="King County") %>% droplevels()
```

`r nrow(data)` trees were included in King County

```{r}
data$top.dieback <- as.factor(data$top.dieback)
data$thinning <- as.factor(data$thinning)
```

# Visualize King County Data

```{r}
#ggplot(data,aes(binary.tree.canopy.symptoms,kc.af.f,fill=binary.tree.canopy.symptoms))+geom_violin()+coord_flip()+theme_bw()+ scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929"))+guides(fill=FALSE)
```

```{r}
p1 <- ggplot(data,aes(binary.tree.canopy.symptoms,fill=binary.tree.canopy.symptoms))+geom_histogram(stat="count",alpha=0.5)+theme_bw()+coord_flip()+ scale_fill_manual(values=c("#7fcdbb","#fe9929"))+guides(fill=FALSE)
```

```{r}
p2 <- ggplot(data,aes(reclassified.tree.canopy.symptoms,fill=reclassified.tree.canopy.symptoms))+geom_histogram(stat="count",alpha=0.5)+theme_bw()+coord_flip()+ scale_fill_manual(values=c("#fe9929","#7fcdbb","#fe9929","#fe9929"))+guides(fill=FALSE)
```

```{r}
p1 + p2
```

```{r}
#p3 <- ggplot(data,aes(binary.tree.canopy.symptoms)) +facet_wrap(~field.optional...tree.size) +geom_histogram(stat="count")+theme_bw()+coord_flip()
#p3
```

```{r}
#p4 <- ggplot(data,aes(binary.tree.canopy.symptoms)) +facet_wrap(~field.optional...site.location.description) +geom_histogram(stat="count")+theme_bw()+coord_flip()
#p4
```

```{r}
p5 <-ggplot(data,aes(binary.tree.canopy.symptoms,kc.af.f, fill=binary.tree.canopy.symptoms ))+geom_boxplot(alpha=0.5)+theme_bw()+coord_flip() + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929"))+guides(fill=FALSE) +labs(x="Tree Condition",y=NULL)
```

```{r}
p6 <-ggplot(data,aes(kc.af.f,fill=binary.tree.canopy.symptoms))+geom_density(alpha=0.5) + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929")) +theme_bw() +labs(x="Afternoon Temperature") +guides(fill=FALSE)
```

```{r}
p5 / p6
```

```{r}
p7 <-ggplot(data,aes(top.dieback,kc.af.f, fill=top.dieback))+geom_boxplot(alpha=0.5)+theme_bw()+coord_flip() + scale_fill_manual(name="Top Dieback",values=c("#7fcdbb","#fe9929"))+guides(fill=FALSE) +labs(x="Top Dieback",y=NULL)
```

```{r}
p8 <-ggplot(data,aes(kc.af.f,fill=top.dieback))+geom_density(alpha=0.5) + scale_fill_manual(name="Top Dieback",values=c("#7fcdbb","#fe9929")) +theme_bw() +labs(x="Afternoon Temperature") +guides(fill=FALSE)
```

```{r}
p7 / p8
```


# Analyses



## Temperature Time Periods

```{r}
binomial.daily <- glmmTMB(binary.tree.canopy.symptoms ~ mean.temp.daily.f ,family=binomial,data=data)
binomial.am <- glmmTMB(binary.tree.canopy.symptoms ~ kc.am.f ,data=data,family=binomial)
binomial.af <- glmmTMB(binary.tree.canopy.symptoms ~ kc.af.f ,data=data,family=binomial)
binomial.pm <- glmmTMB(binary.tree.canopy.symptoms ~ kc.pm.f,data=data,family=binomial)
```

### Best Fit Comparison

```{r}
AICtab(binomial.daily,binomial.am,binomial.af,binomial.pm)
```

> Afternoon heat was the best fit


## Response 1.0: healthy / unhealthy

```{r}
binomial.af <- glmmTMB(binary.tree.canopy.symptoms ~ kc.af.f, data=data,family=binomial)
```

### Summary

```{r}
summary(binomial.af)
```

Tree health is associated with urban heat. The probability of success (tree categorized as unhealthy) increases with urban heat

Unhealthy trees are considered success (1) because it is the second level. The probability of failure (0) is the first level (Healthy).

```{r}
levels(data$binary.tree.canopy.symptoms)
```

The probability of a tree being classified as unhealthy increases with urban temperature. 


### Visualize effects of predictors

First, generate predictions from model

```{r}
# Create a data frame for predictions
new.data <- expand.grid(
  kc.af.f = seq(min(data$kc.af.f), max(data$kc.af.f), length.out = 100)  # keeping predictor2 constant at its mean
)
```

```{r}
new.data$predicted_response <- predict(binomial.af, newdata = new.data, type = "response")
```

```{r}
#head(new.data)
```

```{r}
ggplot(new.data, aes(x = kc.af.f, y = predicted_response)) +
  geom_line() +
  labs(x = "Afternoon Temperature", y = "Predicted Response") +
  theme_bw()
```

The probability of tree being classified as unhealthy increases with afternoon temperature


---

## Response 2.0: healthy / top-dieback

```{r}
data.top <- data %>% filter(reclassified.tree.canopy.symptoms=="Healthy"|reclassified.tree.canopy.symptoms=="Dead Top") %>% droplevels()
```



```{r}
top.dieback.binomial.af <- glmmTMB(top.dieback ~ kc.af.f,data=data.top,family=binomial)
```

### Summary

```{r}
summary(top.dieback.binomial.af)
```


```{r}
levels(data$top.dieback)
```

The probability of a tree having top dieback increases with urban temperature. 

### Visualize effects of predictors

First, generate predictions from model

```{r}
# Create a data frame for predictions
new.data.2 <- expand.grid(
  kc.af.f = seq(min(80), max(150), length.out = 1000)  # keeping predictor2 constant at its mean
)
```


```{r}
#head(new.data.2)
```


```{r}
new.data.2$predicted_response <- predict(top.dieback.binomial.af, newdata = new.data.2, type = "response")
```

```{r}
# Create a data frame for predictions
new.data.2.inf <- expand.grid(
  kc.af.f = seq(min(data.top$kc.af.f), max(150), length.out = 1000)  # keeping predictor2 constant at its mean
)
```

```{r}
new.data.2.inf$predicted_response <- predict(top.dieback.binomial.af, newdata = new.data.2.inf, type = "response")
```


```{r}
p <- ggplot(new.data.2.inf, aes(x = kc.af.f, y = predicted_response)) + geom_line() + labs(x = "Afternoon temperature (F)", y = "Dieback Probability") + scale_x_continuous(limits = c(80, 115)) +theme_minimal()

p
```


### Calculate Dieback Probability Categories


```{r}
#head(new.data.2[new.data.2$predicted_response>0.028 & new.data.2$predicted_response<0.032,])
```

King County redcedar at 84.39 degrees had a 0.03 probability

```{r}
#head(new.data.2[new.data.2$predicted_response>0.024 & new.data.2$predicted_response<0.028,])
```


King County redcedar at 83.96 degrees had a 0.025 probability

```{r}
#head(new.data.2[new.data.2$predicted_response>0.048 & new.data.2$predicted_response<0.055,])
```

King County trees at 85.74 had about a 0.05 probability

```{r}
#head(new.data.2[new.data.2$predicted_response>0.095,])
```

King County redcedar at about 87.56 degrees had a 0.1 predicted response. 


```{r}
#head(new.data.2[new.data.2$predicted_response>0.24 & new.data.2$predicted_response<0.26,])
```

King County redcedar at about 90.29 degrees had a 0.25 predicted response. 


* King County Dieback Probability Categories (25%)
  + Most Concern (>25%): > 90.29
  + Moderate Concern (5-25%): 85.74 - 90.29
  + Lowest Concern (<5%): < 85.74
  
  

```{r}
p <- ggplot(new.data.2, aes(x = kc.af.f, y = predicted_response)) + geom_line() + labs(x = "Afternoon temperature (F)", y = "Dieback Probability") + scale_x_continuous(limits = c(80, 92.5),breaks=c(80,82.5,85,87.5,90,92.5)) + scale_y_continuous(limits = c(0, .3)) +theme_minimal()

p2 <- p + annotate(geom="rect",ymin = -Inf, ymax = Inf, 
              xmin = -Inf, xmax = 85.74, fill = alpha('#DDA0DD',0.5)) +
    annotate(geom="rect",ymin = -Inf, ymax = Inf, 
              xmin = 85.74, xmax = 90.29, fill = alpha('#7fcdbb',0.5)) + 
    annotate(geom="rect",ymin = -Inf, ymax = Inf, 
          xmin = 90.29, xmax = Inf, fill = alpha('#fe9929',0.5))
p3 <- p2 + annotate("text", x=82, y=.15, label= "Lowest\nConcern") + annotate("text", x=87.5, y=.15, label= "Moderate\nConcern") + annotate("text", x=91.9, y=.15, label= "Most\nConcern")
p3
```
  
  
* King County Top-Dieback Probability Categories (10%)
  + Most Concern (>10%): > 87.56
  + Moderate Concern (2.5-10%): 83.96-87.56
  + Lowest Concern (<2.5%): < 83.96

Convert to Celsius 



```{r}
p <- ggplot(new.data.2, aes(x = kc.af.f, y = predicted_response)) + geom_line() + labs(x = "Afternoon temperature (F)", y = "Dieback Probability") + scale_x_continuous(limits = c(80, 92.5),breaks=c(80,82.5,85,87.5,90,92.5)) + scale_y_continuous(limits = c(0, .3)) +theme_minimal()

p2 <- p + annotate(geom="rect",ymin = -Inf, ymax = Inf, 
              xmin = -Inf, xmax = 83.96, fill = alpha('#DDA0DD',0.5)) +
    annotate(geom="rect",ymin = -Inf, ymax = Inf, 
              xmin = 83.96, xmax = 87.56, fill = alpha('#7fcdbb',0.5)) + 
    annotate(geom="rect",ymin = -Inf, ymax = Inf, 
          xmin = 87.56, xmax = Inf, fill = alpha('#fe9929',0.5))
p3 <- p2 + annotate("text", x=82, y=.15, label= "Lowest\nConcern") + annotate("text", x=85.8, y=.15, label= "Moderate\nConcern") + annotate("text", x=90, y=.15, label= "Most\nConcern")
p3
```



---

## Response 3.0: healthy / thinning

```{r}
data.thin <- data %>% filter(reclassified.tree.canopy.symptoms=="Healthy"|reclassified.tree.canopy.symptoms=="Thinning Canopy") %>% droplevels()
```

```{r}
thinning.binomial.af <- glmmTMB(thinning ~ kc.af.f,data=data.thin,family=binomial)
```

### Summary

```{r}
summary(thinning.binomial.af)
```


```{r}
levels(data.thin$thinning)
str(data.thin$thinning)
```

There was no evidence the probability of a tree being classified as thinning is related to afternoon heat.  



### Visualize effects of predictors

Generate predictions from model

```{r}
# Create a data frame for predictions
new.data.3 <- expand.grid(
  kc.af.f = seq(min(data.thin$kc.af.f), max(data.thin$kc.af.f), length.out = 100),  # keeping predictor2 constant at its mean
  Area = 0  # assuming random effect is zero for prediction
)
```

```{r}
new.data.3$predicted_response <- predict(thinning.binomial.af, newdata = new.data.3, type = "response")
```

```{r}
ggplot(new.data.3, aes(x = kc.af.f, y = predicted_response)) +
  geom_line() +
  labs(x = "Distance from Mean Temperature", y = "Dieback Probability") +
  theme_minimal()
```





